Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-10 main.tex
comments

We start from a countable set $\VAR=\set{x,y,z,\ldots}$ of variables,
a countable set $\FUN=\set{f,g,\ldots}$ of function symbols, and a
counrable set $\PRED=\set{p,q,\ldots}$ of predicate symbols. Each function symbol and
each predicate symbol has a finite arity. The grammars below generate
the set $\TERM$ of \bfit{terms}, denoted by $s,t,u,\ldots$, the set
$\ATOM$ of \bfit{atoms}, denoted by $a,b,c,\ldots$, and the set
$\FORM$ of \bfit{formulas}, denoted by $A,B,C,\ldots$:
\begin{equation*}
  \begin{array}{r@{~}l}
    t &\coloneqq~ x \mid f(t_1,\dots,t_n)
    \\[.5ex]
    a &\coloneqq~ \ttt\mid\fff\mid p(t_1,\dots,t_n) \mid \dual p(t_1,\dots,t_n)
    \\[.5ex]
    A &\coloneqq~ a \mid A\vlan A \mid A\vlor A \mid \exists x.A \mid \forall x.A
\end{array}
\end{equation*}
where the arity of $f$ and $p$ is $n$. Note, that in this paper we
consider the truth constants $\ttt$ (\emph{true}) and $\fff$
(\emph{false}) as atoms, and we consider all formulas in negation
normal form. The \bfit{negation} $(\dual\cdot)$ is defined for all
atoms and formulas via the De Morgan laws as follows:
\begin{equation*}
  \begin{array}{cc}
  \dual\ttt =\fff &
  \dual{p(t_1,\ldots,t_n)} = \dual{p}(t_1,\ldots,t_n) \\[.5ex]
  \dual\fff = \ttt &
  \dual{\dual p(t_1,\ldots,t_n)} = p(t_1,\ldots,t_n) \\[.5ex]
  \dual{A \vlan B} = \dual{A} \ \vlor \ \dual{B} &
  \dual{\exists x.A} = \forall x.\dual{A} \\[.5ex]
  \dual{A \vlor B} = \dual{A} \ \vlan \ \dual{B} & 
  \dual{\forall x.A} = \exists x.\dual{A} 
  \end{array}
\end{equation*}
\todo{add missing equations and make it look nice}
\lutz{don't we need   $\dual{\dual a} = a$? (then it should follow that $\dual{\dual A} = A$)}

\juihsuan{isn't it given by the first four equations? I think that we use the
notation of negation for atoms and formulas. We may also need to precise that each
predicate symbol $p$ has a dual $\dual{p}$.}\lutz{we either have to say $\dual{\dual a} = a$ or $\dual{\dual p} = p$. I prefer the first.}

@@

In a graph $\gG$, the \bfit{neighbourhood} $N(v)$ of a vertex $v \in\vG$ is defined as
the set $\set{w \mid vw \in\eG}$. A set $W \subseteq \vG$ induces a \bfit{matching} in
$\gG$ if for all $w \in W$, $N(w) \cap W$ is a singleton.
A \bfit{module} is a set $M \subseteq \vG$ such that $N(v) \ M = N(w) \ M$
for all $v, w \in M$. A module $M$ is \bfit{strong} if for every
module $M'$, we have $M' \subseteq M$, $M \subseteq$ or $M \cap M' =
\emptyset$. A module is \bfit{proper} if it has two or more vertices.

@@

For a formula $A$, we can define its
associated fograph $\graphof(A)$ inductively by: 
\begin{equation*}
  \begin{array}{l}
    \graphof{\alpha} = \cdot \alpha (\text{single-vertex graph whose vertex is 
    labelled by } \alpha); \\[.5ex]
    \graphof{A \cor B} = \graphof{A} + \graphof{B};
    \graphof{A \cand B} = \graphof{A} \times \graphof{B}; \\[.5ex]
    \graphof{\forall x . A} = \cdot x + \graphof{A};
    \graphof{\exists x . A} = \cdot x \times \graphof{A}.
  \end{array}
\end{equation*}
\todo{make look nice, use array} \juihsuan{done}\lutz{use the single macro (see above), so that we can change the notation later if we need to (and make the array with two columns)}

@@

Let $\gG$ be a linked fograph. The set of links can be seen as a unification problem
by identifying dual predicate symbols. A \bfit{dualizer} of $\gG$ is an assignment
unifying all the links of $\gG$. Since a first-order unification problem is either
solvable or has a most general unifier, we can define the notion of \bfit{most
general dualizer}. A \bfit{dependency} is a pair $(\cdot x, \cdot y)$ of an
existential binder $\cdot x$ and a universal binder $\cdot y$ such that the most
general dualizer assigns to $x$ a term containing $y$. A \bfit{leap} is either a
link or a dependency. The \bfit{leap graph} $\lgG$ of $\gG$ is the undirected
graph $(\vG, \lpG)$ where $\lpG$ is the set of leaps of $\gG$. A set $W
\subseteq \vG$ induces a
\bfit{bimatching} in $\gG$ if $W$ induces a matching in $\gG$ and in the leap graph
$\lgG$ as well.

\todo{define matching} \juihsuan{done in the cograph part}\lutz{no, it should be here}

@@

\begin{definition}
A \bfit{first-order combinatorial proof} of a fograph $G$ is a skew
bifibration $f: N \rightarrow G$ where $N$ is a fonet. A \bfit{first-order
combinatorial proof} of a formula $A$ is a combinatorial proof of its graph
$\graphof(A)$.
\end{definition}

>>>>

We start from a countable set $\VAR=\set{x,y,z,\ldots}$ of variables,
a countable set $\FUN=\set{f,g,\ldots}$ of function symbols, and a
counrable set $\PRED=\set{p,q,\ldots}$ of predicate symbols. Each function symbol and
each predicate symbol has a finite arity. The grammars below generate
the set $\TERM$ of \bfit{terms}, denoted by $s,t,u,\ldots$, the set
$\ATOM$ of \bfit{atoms}, denoted by $a,b,c,\ldots$, and the set
$\FORM$ of \bfit{formulas}, denoted by $A,B,C,\ldots$:
\begin{equation*}
  \begin{array}{r@{~}l}
    t &\coloneqq~ x \mid f(t_1,\dots,t_n)
    \\[.5ex]
    a &\coloneqq~ \ttt\mid\fff\mid p(t_1,\dots,t_n) \mid \dual p(t_1,\dots,t_n)
    \\[.5ex]
    A &\coloneqq~ a \mid A\vlan A \mid A\vlor A \mid \exists x.A \mid \forall x.A
\end{array}
\end{equation*}
where the arity of $f$ and $p$ is $n$. Note, that in this paper we
consider the truth constants $\ttt$ (\emph{true}) and $\fff$
(\emph{false}) as atoms, and we consider all formulas in negation
normal form. The \bfit{negation} $(\dual\cdot)$ is defined for all
atoms and formulas via the De Morgan laws as follows:
\begin{equation*}
  \begin{array}{cc}
  \dual\ttt =\fff &
  \dual{p(t_1,\ldots,t_n)} = \dual{p}(t_1,\ldots,t_n) \\[.5ex]
  \dual\fff = \ttt &
  \dual{\dual p(t_1,\ldots,t_n)} = p(t_1,\ldots,t_n) \\[.5ex]
  \dual{\dual a} = a & \dual{\exists x.A} = \forall x.\dual{A} \\[.5ex]
  \dual{A \vlan B} = \dual{A} \ \vlor \ \dual{B} & \dual{\forall x.A} = \exists x.\dual{A} 
  \\[.5ex]
  \dual{A \vlor B} = \dual{A} \ \vlan \ \dual{B} &  
  \end{array}
\end{equation*}
\todo{add missing equations and make it look nice}
\lutz{don't we need   $\dual{\dual a} = a$? (then it should follow that $\dual{\dual A} = A$)}

\juihsuan{isn't it given by the first four equations? I think that we use the
notation of negation for atoms and formulas. We may also need to precise that each
predicate symbol $p$ has a dual $\dual{p}$.}\lutz{we either have to say $\dual{\dual a} = a$ or $\dual{\dual p} = p$. I prefer the first.}
\juihsuan{okay I see. I didn't realize that the negation is regarded as a
connective for predicate symbols. The notation got me confused since $\dual{p}$
is also a predicate symbol.}

@@

In a graph $\gG$, the \bfit{neighbourhood} $N(v)$ of a vertex $v \in\vG$ is defined as
the set $\set{w \mid vw \in\eG}$. A \bfit{module} is a set $M \subseteq \vG$ such that $N(v) \ M = N(w) \ M$
for all $v, w \in M$. A module $M$ is \bfit{strong} if for every
module $M'$, we have $M' \subseteq M$, $M \subseteq$ or $M \cap M' =
\emptyset$. A module is \bfit{proper} if it has two or more vertices.

@@

For a formula $A$, we can define its
associated fograph $\graphof(A)$ inductively by: 
\begin{equation*}
  \begin{array}{cc}
    \graphof{\alpha} = \single \alpha & 
    \graphof{\forall x . A} = \single x + \graphof{A} \\[.5ex]
    \graphof{A \cor B} = \graphof{A} + \graphof{B} &
    \graphof{\exists x . A} = \single x \times \graphof{A} \\[.5ex]
    \graphof{A \cand B} = \graphof{A} \times \graphof{B} &

  \end{array}
\end{equation*}
where $\single \alpha$ (resp. $\single x$) is a single-vertex graph whose vertex is 
    labelled by $\alpha$ (resp. $x$).
\todo{make look nice, use array} \juihsuan{done}\lutz{use the single macro (see
above), so that we can change the notation later if we need to (and make the
array with two columns)} \juihsuan{fixed}

@@

Let $\gG$ be a linked fograph. The set of links can be seen as a unification problem
by identifying dual predicate symbols. A \bfit{dualizer} of $\gG$ is an assignment
unifying all the links of $\gG$. Since a first-order unification problem is either
solvable or has a most general unifier, we can define the notion of \bfit{most
general dualizer}. A \bfit{dependency} is a pair $(\single x, \single y)$ of an
existential binder $\single x$ and a universal binder $\single y$ such that the most
general dualizer assigns to $x$ a term containing $y$. A \bfit{leap} is either a
link or a dependency. The \bfit{leap graph} $\lgG$ of $\gG$ is the undirected
graph $(\vG, \lpG)$ where $\lpG$ is the set of leaps of $\gG$. A set $W \subseteq \vG$ induces a \bfit{matching} in
$\gG$ if for all $w \in W$, $N(w) \cap W$ is a singleton, and induces a
\bfit{bimatching} in $\gG$ if it induces a matching in $\gG$ and in the leap graph
$\lgG$ as well.

\todo{define matching} \juihsuan{done in the cograph part}\lutz{no, it should be here}
\juihsuan{fixed}

@@

\begin{definition}
A \bfit{first-order combinatorial proof} of a fograph $G$ is a skew
bifibration $f: N \rightarrow G$ where $N$ is a fonet. A \bfit{first-order
combinatorial proof} of a formula $A$ is a combinatorial proof of its graph
$\graphof{A}$.
\end{definition}
